# Statistical Analyses

```{r}

# Load libraries
library(tidyr)
library(dplyr)

# Load data and add IDs
data_path <- "microvolunteering.csv"
data <- read.csv(data_path, header = TRUE)
data$Participant <- seq_len(nrow(data)) 
```

## Inferential Statistical Tests

```{r}
### Wilcoxon Ranked-Sign Test ### 

happy_critical_test <- wilcox.test(
  subset(data, Task == "Critical")$Happy.POST,
  subset(data, Task == "Critical")$Happy.PRE,
  paired = TRUE,
  exact = FALSE
)
happy_noncritical_test <- wilcox.test(
  subset(data, Task == "Non-critical")$Happy.POST,
  subset(data, Task == "Non-critical")$Happy.PRE,
  paired = TRUE,
  exact = FALSE
)
stim_critical_test <- wilcox.test(
  subset(data, Task == "Critical")$Stimulated.POST,
  subset(data, Task == "Critical")$Stimulated.PRE,
  paired = TRUE,
  exact = FALSE
)
stim_noncritical_test <- wilcox.test(
  subset(data, Task == "Non-critical")$Stimulated.POST,
  subset(data, Task == "Non-critical")$Stimulated.PRE,
  paired = TRUE,
  exact = FALSE
)
controlled_critical_test <- wilcox.test(
  subset(data, Task == "Critical")$Controlled.POST,
  subset(data, Task == "Critical")$Controlled.PRE,
  paired = TRUE,
  exact = FALSE
)
controlled_noncritical_test <- wilcox.test(
  subset(data, Task == "Non-critical")$Controlled.POST,
  subset(data, Task == "Non-critical")$Controlled.PRE,
  paired = TRUE,
  exact = FALSE
)

### Mann-Whitney U Test ### 

# Compute (post - pre) mood changes for each of the three affect dimensions 
# and add a column for each to the data frame
data <- data %>%
  mutate(
    Happy.DIFF = Happy.POST - Happy.PRE,
    Stimulated.DIFF = Stimulated.POST - Stimulated.PRE,
    Controlled.DIFF = Controlled.POST - Controlled.PRE
  )

# Create data frames for "critical" and "non-critical" groups
noncritical_group <- subset(data, Task == "Non-critical")
critical_group <- subset(data, Task == "Critical")

# Perform Mann-Whitney U tests for changes in each affect dimension
happy_test <- wilcox.test(
  noncritical_group$Happy.DIFF, 
  critical_group$Happy.DIFF, 
  paired = FALSE,
  exact = FALSE
)
stimulated_test <- wilcox.test(
  noncritical_group$Stimulated.DIFF,
  critical_group$Stimulated.DIFF,
  paired = FALSE,
  exact = FALSE
)
controlled_test <- wilcox.test(
  noncritical_group$Controlled.DIFF,
  critical_group$Controlled.DIFF,
  paired = FALSE,
  exact = FALSE
)
```

### WILCOXON RANKED-SIGN TEST RESULTS

Happy Critical Test
```{r}
print(happy_critical_test)
```

Happy Non-critical Test
```{r}
print(happy_noncritical_test)
```

Stimulated Critical Test

```{r}
print(stim_critical_test)
```

Stimulated Non-critical Test

```{r}
print(stim_noncritical_test)
```

Controlled Critical Test

```{r}
print(controlled_critical_test)
```

Controlled Non-critical Test

```{r}
print(controlled_noncritical_test)
```

### MANN-WHITNEY U TEST RESULTS

Happy Test

```{r}
print(happy_test)
```

Stimulated Test

```{r}
print(stimulated_test)
```

Controlled Test

```{r}
print(controlled_test)
```


## Descriptive Statistics

```{r}
# Create vectors to store values for table creation
critical_pre_post <- c()
noncritical_pre_post <- c()
critical_diff <- c()
noncritical_diff <- c()
```


### Mean

Happy

```{r}
happy_critical_diff_mean <- 
  mean(subset(data, Task == "Critical")$Happy.POST) - mean(subset(data, Task == "Critical")$Happy.PRE)
print(happy_critical_diff_mean)

happy_noncritical_diff_mean <- 
  mean(subset(data, Task == "Non-critical")$Happy.POST) - mean(subset(data, Task == "Non-critical")$Happy.PRE)
print(happy_noncritical_diff_mean)

# Assign values to vectors to create table
critical_pre_post[1] = mean(subset(data, Task == "Critical")$Happy.PRE)
critical_pre_post[4] = mean(subset(data, Task == "Critical")$Happy.POST)
noncritical_pre_post[1] = mean(subset(data, Task == "Non-critical")$Happy.PRE)
noncritical_pre_post[4] = mean(subset(data, Task == "Non-critical")$Happy.POST)
critical_diff[1] = happy_critical_diff_mean
noncritical_diff[1] = happy_noncritical_diff_mean
```

Stimulated

```{r}
stim_critical_diff_mean <- 
  mean(subset(data, Task == "Critical")$Stimulated.POST) - mean(subset(data, Task == "Critical")$Stimulated.PRE)
print(stim_critical_diff_mean)

stim_noncritical_diff_mean <- 
  mean(subset(data, Task == "Non-critical")$Stimulated.POST) - mean(subset(data, Task == "Non-critical")$Stimulated.PRE)
print(stim_noncritical_diff_mean)

# Assign values to vectors to create table
critical_pre_post[2] = mean(subset(data, Task == "Critical")$Stimulated.PRE)
critical_pre_post[5] = mean(subset(data, Task == "Critical")$Stimulated.POST)
noncritical_pre_post[2] = mean(subset(data, Task == "Non-critical")$Stimulated.PRE)
noncritical_pre_post[5] = mean(subset(data, Task == "Non-critical")$Stimulated.POST)
critical_diff[2] = stim_critical_diff_mean
noncritical_diff[2] = stim_noncritical_diff_mean
```

Controlled

```{r}
cont_critical_diff_mean <- 
  mean(subset(data, Task == "Critical")$Controlled.POST) - mean(subset(data, Task == "Critical")$Controlled.PRE)
print(cont_critical_diff_mean)

cont_noncritical_diff_mean <- 
  mean(subset(data, Task == "Non-critical")$Controlled.POST) - mean(subset(data, Task == "Non-critical")$Controlled.PRE)
print(cont_noncritical_diff_mean)

# Assign values to vectors to create table
critical_pre_post[3] = mean(subset(data, Task == "Critical")$Controlled.PRE)
critical_pre_post[6] = mean(subset(data, Task == "Critical")$Controlled.POST)
noncritical_pre_post[3] = mean(subset(data, Task == "Non-critical")$Controlled.PRE)
noncritical_pre_post[6] = mean(subset(data, Task == "Non-critical")$Controlled.POST)
critical_diff[3] = cont_critical_diff_mean
noncritical_diff[3] = cont_noncritical_diff_mean
```

```{r}
## Create table to display these values
mean_sam <- data.frame(critical_pre_post, noncritical_pre_post)
row.names(mean_sam) <- c('Happy.PRE', 'Stimulated.PRE', 'Dominance.PRE', 'Happy.POST', 'Stimulated.POST', 'Dominance.POST')
mean_sam

mean_diff_sam <- data.frame(critical_diff, noncritical_diff)
row.names(mean_diff_sam) <- c('Happy', 'Stimulated', 'Dominance')
mean_diff_sam
```

### Visualize changes 

```{r}

# Citation: I generated the visualization code below with GPT 4.5

# Load ggplot2 library
library(ggplot2)

# Create a data frame for the plot using your calculated mean differences
plot_data <- data.frame(
  Dimension = rep(c("Valence", "Arousal", "Dominance"), each = 2),
  Task = rep(c("Critical Task", "Non-critical Task"), times = 3),
  Value = c(
    happy_critical_diff_mean, happy_noncritical_diff_mean,
    stim_critical_diff_mean, stim_noncritical_diff_mean,
    cont_critical_diff_mean, cont_noncritical_diff_mean
  )
)

# Generate the grouped bar chart
ggplot(plot_data, aes(x = Dimension, y = Value, fill = Task)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.6) +
  scale_fill_manual(values = c("Critical Task" = "#FF9999", "Non-critical Task" = "#99CCFF")) +
  theme_minimal() +
  labs(
    title = "Mean Change in Emotional States by Task Type",
    x = "Dimension",
    y = "Mean Change (Post - Pre)",
    fill = "Task Type"
  ) +
  ylim(-1, 1) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  theme(
    text = element_text(size = 14),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```


### Median

Happy

```{r}
happy_critical_diff_median <- 
  median(subset(data, Task == "Critical")$Happy.POST) - median(subset(data, Task == "Critical")$Happy.PRE)
print(happy_critical_diff_median)

happy_noncritical_diff_median <- 
  median(subset(data, Task == "Non-critical")$Happy.POST) - median(subset(data, Task == "Non-critical")$Happy.PRE)
print(happy_noncritical_diff_median)
```

Stimulated

```{r}
stim_critical_diff_median <- 
  median(subset(data, Task == "Critical")$Stimulated.POST) - median(subset(data, Task == "Critical")$Stimulated.PRE)
print(stim_critical_diff_median)

stim_noncritical_diff_median <- 
  median(subset(data, Task == "Non-critical")$Stimulated.POST) - median(subset(data, Task == "Non-critical")$Stimulated.PRE)
print(stim_noncritical_diff_median)
```

Controlled

```{r}
cont_critical_diff_median <- 
  median(subset(data, Task == "Critical")$Controlled.POST) - median(subset(data, Task == "Critical")$Controlled.PRE)
print(cont_critical_diff_median)

cont_noncritical_diff_median <- 
  median(subset(data, Task == "Non-critical")$Controlled.POST) - median(subset(data, Task == "Non-critical")$Controlled.PRE)
print(cont_noncritical_diff_median)
```